NumLin: Linear Types for Linear Algebra

### 2018-11-12 write-up/paper/implementation.tex
Start Performance Metrics

Because all functions must have their argument types explicitly annotated,
inferring the correct fraction at a call-site is simply a matter of
unification. We strongly suspect that \emph{full-inference of fractional
permissions is similarly just matter of unification} (thanks to an
experimental implementation of just this feature), even though the formal
system we present here is for an explicitly-typed language.

@@

Here, evaluate the performance of the examples from the second
section.  Compare with your C implementations, and perhaps as well as
the straightforward math transcribed into (Matlab/R/Numpy?).

@@

>>>>

Because all functions must have their argument types explicitly annotated,
inferring the correct fraction at a call-site is simply a matter of
unification. We believe \emph{full-inference of fractional permissions is
similarly just matter of unification} (thanks to an experimental implementation
of just this feature), even though the formal system we present here is for an
explicitly-typed language.

@@

% Here, evaluate the performance of the examples from the second
% section. Compare with your C implementations, and perhaps as well as
% the straightforward math transcribed into (Matlab/R/Numpy?).

We think that using \lang has two primary benefits: safety and performance. We
discuss safety in \ref{subsec:finding_bugs}, where we describe how we used
\lang\ to find linearity and aliasing bugs in a linear algebra algorithm that
was \emph{generated} by another program.

For performance, we measured the execution four of three equivalent
implementations of a Kalman filter: in C (using \textsc{Cblas}), \lang\ (using
\textsc{Owl}'s low-level \textsc{Cblas} bindings), OCaml (using \textsc{Owl}'s
intended, safe/copying-by-default interface), and Python (using
\textsc{NumPy}).  Measurements were made in in micro-seconds, against an
exponentially (powers of 5) increasing scaling factor for matrix size
parameters $n=5$ and $k=3$. The results in Figure~\ref{fig:timings} are as we
expected: C is the fastest, followed by \lang, with OCaml and Python last.

We expected the C implementation to be faster than the \lang\ one because the
latter has the additional (but relatively low) overhead of dimension checks and
crossing the OCaml/C FFI for each call to a \textsc{Cblas} routine, even though
the calls and their order are exactly the same. We expected the OCaml and
Python implementations to be slower because of the unnecessary temporaries they
allocate.

% Need to re-run timings
Differences is timings are quite pronounced at small matrix sizes, but are
still significant at larger ones: for $n=625$, \textsc{Cblas} took $112 \pm
35\, ms$, \lang\ took $105 \pm 25 \, ms$, \textsc{Owl} took $124 \pm 38\, ms$
and \textsc{NumPy} took $112 \pm 12\, ms$;  for $n=3125$, \textsc{Cblas} took
$10.8 \pm 0.7\, s$, \lang\ took $12.0 \pm 1.2 \, s$, \textsc{Owl} took $13.3
\pm 0.2\, s$ and \textsc{NumPy} took $12.7 \pm 0.6\, s$.

@@

% Talk about memory and temporaries here

% Caveats - set-up problems

### 2018-11-12 write-up/paper/intro.tex
Start Performance Metrics

\lang\ allows a novice to understand and work with complicated linear algebra
library APIs, as well as point out subtle aliasing bugs and reduce memory usage
in existing programs. In fact, we were able to use \lang\ to find linearity and
aliasing bugs in a linear algebra program that was \emph{generated} by another
program \emph{specifically designed to translate matrix expressions into an
efficient sequence of calls linear algebra routines}. We were also able to
reduce the number of temporaries used by the same program, using \lang's type
system to guide us.

>>>>

\lang\ allows a novice to understand and work with complicated linear algebra
library APIs, as well as point out subtle aliasing bugs and reduce memory usage
in existing programs. In fact, we were able to use \lang\ to find linearity and
aliasing bugs in a linear algebra algorithm that was \emph{generated} by
another program \emph{specifically designed to translate matrix expressions
into an efficient sequence of calls to linear algebra routines}. We were also
able to reduce the number of temporaries used by the same algorithm, using
\lang's type system to guide us.
